---
title: "IE 582, Homework 2"
author: "Ayberk AKGÜN"
date: "10/31/2019"
output: html_document
---

The libraries I will use:

```{r cars, message=FALSE, warning=FALSE}
library(data.table)
library(dplyr)
library(jpeg)
library(imager)
library(factoextra)
```

I read the data and apply PCA:

```{r pressure}
raw_data <- read.csv("/Users/ayberkakgun/Desktop/Musk1.csv") %>% as.data.frame()
features<-scale(raw_data[3:168])
labels<-raw_data[1:2]
colnames(labels)<-c("BagClass","BagId")
pca<-princomp(features)
str(pca)
eigenvalue<-get_eigenvalue(pca)
head(eigenvalue,10)
```

#a
First 5 components explain more than 70 % of the variability and cumulative contribution gets smaller.

```{r}
eig<-eigenvalue$variance.percent[eigenvalue$cumulative.variance.percent<85]
l<-length(eig)
ggplot()+
    geom_col(aes(x=1:l,y=eig))+
    geom_line(aes(x=1:l,y=eig,col="red"))

```

If we check plot first two components we see the two dimensions are not enough to distinguish bag labels:

```{r}
plot(x=pca$scores[,1],y=pca$scores[,2],col=raw_data$X1+1)
abline(h=0,v=0,lty=3)
```

I calculate distance matrix and perform MDS on it, without forgetting to scale the data.

```{r}
distanceMat<-dist(features,upper=T) %>% as.matrix()
mds<-cmdscale(distanceMat)
plot(mds[,1],mds[,2],
     # main='Location',
     xlab='', ylab='',
    col=raw_data$X1+1)
```

#b 
Once again perform PCA this time on aggregated data,

```{r}
data_agg<-aggregate(raw_data,by=list(raw_data$X1.1),FUN=mean)
features_agg<-scale(data_agg[4:169])
pca_agg<-prcomp(features_agg)
str(pca_agg)
eigenvalue_agg<-get_eigenvalue(pca_agg)
head(eigenvalue_agg,10)
plot(x=pca_agg$x[,1],y=pca_agg$x[,2],col=data_agg$X1+1)
```

Same MDS operation on aggregated data,

```{r}
distanceMat_agg<-dist(features_agg,upper=T) %>% as.matrix()
mds_agg<-cmdscale(distanceMat_agg)
plot(mds_agg[,1],mds_agg[,2],
     # main='Location',
     xlab='', ylab='',
     col=data_agg$X1+1)
```

#c



#Task 2

I read my image, and plot it.

```{r}
img <- readJPEG("/Users/ayberkakgun/Desktop/profil2.jpg")
plot(c(0, 256), c(0, 256), type = "n", xlab = "", ylab = "")
rasterImage(img, 0, 0, 256, 256)
```

I split the channels and plot them seperately.

```{r}
r<-img[1:256,1:256,1]
g<-img[1:256,1:256,2]
b<-img[1:256,1:256,3]
red_palette <- colorRampPalette(c("black","red"))
blue_palette <- colorRampPalette(c("black","blue"))
green_palette <- colorRampPalette(c("black","green")) 
par(mfrow=c(1,3))
# red_palet
image(r,col=red_palette(256),main="red",axes = F)
image(g,col=green_palette(256),main="green",axes=F)
image(b,col=blue_palette(256),main="blue",axes=F)
```

I add noise to my image.

```{r}
minx<-min(img)
maxx<-max(img)*0.1
noise<-matrix(runif(65536,min=minx,max=maxx),256)
r2<-r+noise
r2[r2>1]=1
g2<-g+noise
g2[g2>1]=1
b2<-b+noise
b2[b2>1]=1
# img2 <- rgb(r2, g2, b2)
img2<-img
img2[,,1]<-r2
img2[,,2]<-g2
img2[,,3]<-b2
plot(c(0, 256), c(0, 256), type = "n", xlab = "", ylab = "")
rasterImage(img2, 0, 0, 256, 256)
```

I construct the patches and perform PCA on them.

```{r}
imgr <- load.image("/Users/ayberkakgun/Desktop/profil2.jpg")
imgrr<-grayscale(imgr)
vectors <-matrix(nrow=53824,ncol = 625)
k<-1
for (i in 1:232){
    for(j in 1:232){
        vectors[k,]<-imgrr[i:(i+24),j:(j+24)] %>% as.vector()
        k<-k+1
    }
    }

# pca_img<-princomp(vectors)
```




